A statistically consistent approach to chemical tagging

### 2018-11-13 article/ms.tex
words

\shorttitle{MCFA}
\shortauthors{Casey et al.}

@@

\title{MCFA on Galah}

@@

\begin{abstract}
foo
\end{abstract}

@@

@@

We use the K-means++ algorithm to initially assign data points
randomly to $K$ clusters (in data space). This provides a so-called
hard initial association where each data point is wholly assigned
to a single component. Specifically we calculate the $n \times K$
responsibility matrix $\vec\tau$, where a entry of 1 indicates the
$n$th data point is associated with the $k$th cluster, and an
entry of 0 indicates no association such that
\begin{equation}
	\weight_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk} \quad .
\end{equation}
We then chose 1,000
random data points and factorize that matrix using singular-value
decomposition. We take the principal $p$ eigenvectors as the 
initial guess for the factor loads $\factorloads$, and calculate
an initial estimate of the factor score means
\begin{equation}
	\scoremeans_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk}\vecdata_{n}\factorloads
\end{equation}
\noindent{}and the covariance matrices of the clusters in latent space
\begin{equation}
	\scorecovs_k = \textrm{cov}{\left(\tau_{nk}\vecdata_{n}\right)} \quad .
\end{equation}

@@

\subsection{Toy model with generated data} \label{sec:experiment-toy-model}

@@

\subsection{GALAH} \label{sec:experiment-galah}

The \Galah\ survey is a stellar spectroscopic survey of up to
$10^6$ stars. Up to 30 chemical abundances are expected to be measured
and reported in the final \Galah\ data release.
\todo{more about galah}

We use the second \Galah\ data release \citep{Buder:2018a} which
includes up to \todo{X} chemical abundances measured for \todo{Y}
stars. We selected stars with \texttt{flag\_cannon = 0} to exclude
stars where there is reason to suspect that the stellar parameters
(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
detailed chemical abundances are untrustworthy. For this subset
we examined the measured abundances of \todo{X, etc} and
counted the number of stars with \texttt{flag\_x\_fe = 0}
for each abundance. This flag indicates that there is no known
reason to suspect that the measured chemical abundance is
unreliable.

We sorted the elemental abundances by the number of stars
with reliable measurements, from highest to lowest. Starting
with Fe, we iterated over each element and counted the number
of stars with reliable measurements of that element and all
other elements earlier in the list. We illustrate this process
in Figure~\ref{fig:galah-abundances}, where we show that
if we include the top 15 elemental abundances, we would have
\todo{X} stars with reliable measurements in all fifteen elements.
This is relevant to the approach we use here, as the method
described here requires no missing data.

@@

% Where do we go down to?
% What elements do we add /remove.
\todo{Where do we go down to? What elements do we exclude?}
This leaves us with a sample of \todo{X} stars each with
reliable stellar parameters ($\teff$, $\logg$) and 
\todo{Y} (reliable) measured abundances: \todo{Mn, Fe}.
These elements are produced from multiple nucleosynthetic
pathways, including \todo{X}. 
It is this subsample that we will use for experiments
in chemical tagging.

@@

We trial the number of latent factors from $J = 1$ to $J = 10$, and trial the
number of clustering components from $K = 1$ to $K = 100$. We re-initialise the
mixture for each new trial of $\{J, K\}$, and do not make use of results from
`nearby' mixtures.

% Run this.

% Unknown number of latent factors and unknown number of latent scores

>>>>

\shorttitle{Statistically consistent chemical tagging}
\shortauthors{Casey et al.}

@@

\title{A statistically consistent approach to chemical tagging}

@@

\begin{abstract}
\end{abstract}

@@

\begin{itemize}
	\item Chemical tagging
	\item Limits on the efficacy of chemical tagging
	\item Clustering vs common traits
\end{itemize}

The \Galah\ survey \citep{DaSilva:2015} is a stellar spectroscopic survey 
that uses the High Efficiency and Resolution Multi-Element 
Spectrograph \citep[HERMES][]{Sheinis:2016} on the Australian Astronomical Telescope (AAT).
The survey will observe up to $10^6$ stars in the Milky Way, and
measure up to 30 chemical abundances for each star. This includes
light odd-Z elements (e.g., Na, K), elements produced through
alpha-particle capture (e.g., Mg, Ca, Ti), and elements produced
through the slow (e.g., Ba) and rapid neutron-capture process
(e.g., Eu). These data provide an unparalleled view on the production
of chemical elements in the Milky Way.

@@

We initialise the entries in the factor loads $\factorloads$ by drawing
from a uniform distribution $\factorloads \sim \mathcal{U}\left(-1, +1\right)$.
We assign each data point as belonging to one of the $K$ components
by drawing from a multinomial distribution. Given these assignments and
factor loads, we then estimate the relative weights $\vec\pi$, the 
mean factor scores of each component $\scoremeans$, and the covariance
matrix of factor scores of each component $\scorecovs$.

\begin{comment}
We use the K-means++ algorithm to initially assign data points
randomly to $K$ clusters (in data space). This provides a so-called
hard initial association where each data point is wholly assigned
to a single component. Specifically we calculate the $n \times K$
responsibility matrix $\vec\tau$, where a entry of 1 indicates the
$n$th data point is associated with the $k$th cluster, and an
entry of 0 indicates no association such that
\begin{equation}
	\weight_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk} \quad .
\end{equation}
We then chose 1,000
random data points and factorize that matrix using singular-value
decomposition. We take the principal $p$ eigenvectors as the 
initial guess for the factor loads $\factorloads$, and calculate
an initial estimate of the factor score means
\begin{equation}
	\scoremeans_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk}\vecdata_{n}\factorloads
\end{equation}
\noindent{}and the covariance matrices of the clusters in latent space
\begin{equation}
	\scorecovs_k = \textrm{cov}{\left(\tau_{nk}\vecdata_{n}\right)} \quad .
\end{equation}
\end{comment}

@@

\subsection{Experiment 1: Toy model with generated data} \label{sec:experiment-toy-model}

@@

\subsection{Experiment 2: A subset of 14 chemical abundances in \Galah}
\label{sec:experiment-galah}

We use the second \Galah\ data release \citep{Buder:2018a} which
includes up to 23 chemical abundances measured for 342,682
stars. It is not practical to use all of these measurements
when performing chemical tagging. For example, our experiments
with \Galah\ data do not include lithium abundances because the
photospheric lithium varies throughout a star's lifetime. 
We first selected stars with \texttt{flag\_cannon = 0} to exclude
stars where there is reason to suspect that the stellar parameters
(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
detailed chemical abundances are untrustworthy.

We included the following fourteen chemical abundances: Na, Mg, 
Si, Ca, Sc, Ti, Cr, Mn, Fe, Ni, Cu, Zn, Ba, and Eu.
By requiring that each star has \texttt{flag\_cannon = 0} and
\texttt{flag\_<x> = 0} (where \texttt{<x>} is the relevant label)
for each chemical abundance, this leaves
us with a sample of 12,798 stars each with reliable stellar
parameters ($\teff$, $\logg$) and chemical abundances.
This is relevant to the approach we use here, as the method
described here requires no missing data.
These elements are produced from multiple nucleosynthetic
pathways, and it is this subsample that we will use for 
our initial chemical experiment with \Galah\ data.

\begin{comment}
For this subset
we examined the measured abundances and
counted the number of stars with \texttt{flag\_x\_fe = 0}
for each abundance. This flag indicates that there is no known
reason to suspect that the measured chemical abundance is
unreliable.

We sorted the elemental abundances by the number of stars
with reliable measurements, from highest to lowest. Starting
with Fe, we iterated over each element and counted the number
of stars with reliable measurements of that element and all
other elements earlier in the list. We illustrate this process
in Figure~\ref{fig:galah-abundances}, where we show that
if we include the top 14 elemental abundances, we would have
12,798 stars with reliable measurements in all fourteen elements.
This is relevant to the approach we use here, as the method
described here requires no missing data.
\end{comment}

@@

\begin{comment}
This leaves us with a sample of 12,798 stars each with
reliable stellar parameters ($\teff$, $\logg$) and fourteen
(reliably) measured abundances: Na, Mg, Si, Ca, Sc, Ti, Cr, Mn, 
Fe, Ni, Cu, Zn, Ba, and Eu
\todo{Y} (reliable) measured abundances: \todo{Mn, Fe}.
These elements are produced from multiple nucleosynthetic
pathways, including \todo{X}. 
It is this subsample that we will use for experiments
in chemical tagging.
\end{comment}

@@

We trial the number of latent factors from $J = 1$ to $J = 12$, and trial the
number of clustering components from $K = 1$ to $K = 40$. We re-initialise the
mixture for each new trial of $\{J, K\}$, and do not make use of results from
`nearby' mixtures. After the E-M threshold of $10^{-5}$ has been reached, we
record the log likelihood of the data given the model parameters, the
Bayesian Information Criterion \citep{bic}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:bic}
\end{equation} 
\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
\begin{equation}
 \textrm{pseudo-BIC} = 6(1 + \gamma)\omega{}Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:pseudo-bic}
\end{equation}
\noindent{}where $Q$ is the total number of model parameters in both equations,
and the unknown constants $\omega$ and $\gamma$ have constraints
$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.

\todo{Discuss grid search, inferred factor loads, cluster variances, etc.}

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/galah-experiment-2-gridsearch-ll-contours.png}
	\includegraphics[width=0.45\textwidth]{experiments/galah-experiment-2-gridsearch-bic-contours.png}
	\includegraphics[width=0.45\textwidth]{experiments/galah-experiment-2-gridsearch-pseudobic-contours.png}
    \caption{Model performance metrics for Experiment 2 using a subset of 14 
    		 \Galah\ chemical abundances measured from 12,798 stars. The top 
		 	 panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, the middle panel shows the Bayesian Information
			 Criterion (Equation \ref{eq:bic}), and the lower panel shows
			 the pseudo-Bayesian Information Criterion 
			 (Equation~\ref{eq:pseudo-bic}). The white marker indicates the 
			 lowest value in each panel.}
    \label{fig:experiment-2-gridsearch}
\end{figure}

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-1-gridsearch-factors-wrt-J-and-K.png}
	\caption{Latent factor loads inferred for models with increasing numbers
			 of latent factors and increasing numbers of cluster components
			 in Experiment~2. The latent factors have each been rotated by a
			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
			 that the estimated latent factors in each model are ordered 
			 and oriented to be as similar as the latent factors estimated from
			 the model with only one latent factor. Note that the rotation
			 matrix contains only entries $\{-1, 0, +1\}$. The latent
			 factors for each model shown here were initialised using singular value
			 decomposition on a random subset of the data.
			 }
	\label{fig:experiment-2-gridsearch-factors-init-svd}
\end{figure*}

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-2-gridsearch-factors-wrt-J-and-K.png}
	\caption{Latent factor loads inferred for models with increasing numbers
			 of latent factors and increasing numbers of cluster components
			 in Experiment~2. The latent factors have each been rotated by a
			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
			 that the estimated latent factors in each model are ordered 
			 and oriented to be as similar as the latent factors estimated from
			 the model with only one latent factor. Note that the rotation
			 matrix contains only entries $\{-1, 0, +1\}$. The latent
			 factors for each model shown here were initialised randomly.
			 }
	\label{fig:experiment-2-gridsearch-factors-init-svd}
\end{figure*}

\subsection{Experiment 4: A minimal set of nucleosynthetic trace elements in \Galah}

In this experiment we investigate the effects of initialising the model parameters.
Here we make use of a minimal set of nucleosynthetic trace elements
using the second \Galah\ data release \citep{Buder:2018}. Specifically, we use
reliable chemical abundance measurements of Na (an odd-Z element), Mg and Ca (alpha-elements),
Ni, Fe, and Cr (iron-peak elements), Ba and Y (s-process tracers), and Eu (an r-process tracer).
We assume that there are four latent factors ($J = 4$) and four components ($K = 4$),
and we initialise the latent factors and component parameters in different combinations.
The latent factors can be initialised in four different ways: (1) with entries drawn randomly from $\mathcal{U}\left(-1, +1\right)$, (2) with noisey draws $\mathcal{N}\left(0, 10^{-2}\right)$, (3) using singular value decomposition, and (4) using astrophysical insight to initialise the factor loads. 
\todo{When using astrophysical insight...}
The component parameters are either initialised randomly, or using K-means++ algorithm in
the latent space. For each combination of initialisation procedure, we run 30 optimisations
and record the log likelihood once the E-M procedure has reached a threshold of $10^{-8}$.

The results of this experiment are shown in Figure~\ref{fig:experiment-4-log-L}.
